Depression and suicidality as credible signals of need

### 2021-05-17 Paper.rmd
Minor edits to intro and SI

---
title: "Depression and suicidality as credible signals of need"
author: "Michael R. Gaffney, Kai Adams, Kristen L. Syme, and Edward H. Hagen"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: true
output: 
  bookdown::html_document2:
    number_sections: false
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

Major depression and suicide are generally viewed as pathological responses to stress that elicit aversive responses from others. Alternatively, depression and suicide attempts might involve honest signals of need in response to stress and help to bargain for increased support from reluctant others. To address this hypothesis, we examined responses to emotional signals in an experimental vignette study involving claims of substantial need in the presence of conflicts of interest and private information about whether the signaler was actually in need. In our sample of 1,240 participants recruited from Amazon Mechanical Turk, costlier signals like depression and suicide attempts resulted in greater perceptions of need and increased likelihood of support compared to simple verbal requests and crying without further symptoms. Additionally, the effect of signal type on likelihood of support was largely mediated by the effect of signaling on belief in need as predicted. Our results suggest that depression and suicide attempts, apparent human universals, are relatively successful ways of signaling honest need and eliciting support, consistent with costly signaling explanations.

@@

With a cooperative species like our own, the ubiquity of conflicts of interests means that there will always be disagreement over the levels of investment in a cooperative endeavor and the division of the resulting benefits. In many such cases, individuals who were dissatisfied with the terms of cooperation could seek other partners [@hammerstein2016]. In many other cases, however, it would be difficult or impossible to switch partners. Spouses who were dissatisfied with their partner's investment in their new infant, for example, would not easily find a different partner to invest in that infant [@Hagen:1999]. Similarly, an adolescent who was dissatisfied with her parent's investment in her would not easily find other parents who were willing to invest more, nor could parents easily produce another adolescent. In these examples, and many cooperative endeavors central to human biological fitness, each party has monopoly power over the benefits they bring to the endeavor -- no one is easily replaced [@Hagen2003].

@@

Formal models of bargaining highlight the role of private information regarding the value each party places the cooperative endeavor. Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay* -- the more one values the endeavor, the greater the cost of delay. In a classic non-cooperative game theory model of bargaining over the division of benefits, @rubinstein1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay. If valuations are private information, however, then immediate agreement might not be possible because each party has an incentive to request an amount greater than their actual valuation, which would harm the other party. A willingness to delay is a credible signal that one has a low valuation of the endeavor, and is therefore genuinely in need, because for those with a low valuation the benefit of waiting for a better offer outweighs the low cost of delay whereas for those for those with a high valuation the benefit of waiting for a better offer does not outweigh the higher cost of delay. Once valuations are known, the game reduces to the one in @rubinstein1982, and the parties can reach an agreement on the division of benefits [@KennanWilson1993; @feinberg2005]. @Hagen2003 proposed that the substantial reduction in productivity in depression corresponds to a willingness to delay, and is therefore a credible signal of need. In other words, reduced productivity is relatively less costly for those whose efforts are currently not yielding many fitness-relevant benefits than for those whose efforts are yielding substantial benefits [@Hagen2003].

@@

Suicidality is closely linked to depression. Suicidal ideation is one of nine diagnostic criteria for a major depressive episode [MDE, @dsm5], it is associated with depression across cultures [@haroz2017], and depression is a major risk factor for suicidal behavior [@hawton2013].

Anthropology has long viewed suicidality as largely the result of social problems. Early in the field's history, anthropologists tackled the problem of suicide as manifested in the small, traditional societies in which they conducted their fieldwork. Suicide, they found, was commonly a form of protest, revenge, and/or appeal (e.g., Malinowski, 1932; Niehaus, 2012). Some ethnographers, such as Hezel (1987), conceptualized suicide as a form of anger, whereas others, such as Giddens (1964), conceptualized suicide as a form of social pressure. Counts (1980) emphasized the powerlessness of suicide victims.

Common to almost all theoretical and empirical investigations of suicide is a focus on *completed* suicides, i.e., suicide deaths. The vast majority of suicidal behavior, however, does not result in death. In young adult women in the US, for example, there are hundreds of attempts for every death (see Figure \@ref(fig:cdc)). @SymeEtAl2016 therefore argued that the scientific focus should be on suicide ideation and suicide attempts.

Raymond Firth was one of the first to view suicidality as a gamble to improve one's circumstances in the here and now. Based on observations that suicide attempts often followed loss or conflict and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, or immediate reintegration into the community following negative events (Firth 1936; 1961).

@Rosenthal1993 proposed a credible signaling model of what he, like Firth, viewed as 'gambling-type' suicide attempts. In the Rosenthal model, as long as that risk is greater than or equal to the net benefit a non-needy person would realize by sending the signal (discounted by the risk of death), social partners can trust that signal senders are genuinely in need despite conflicts of interest [@Rosenthal1993]. In the bargaining framework, suicidality could be conceptualized as putting future fitness benefits at risk -- riskier attempts would credibly indicate lower valuation of the future. On these views, most suicides deaths, especially in young, physically healthy individuals, would therefore be the inevitable consequence that some individuals lose this gamble. These and similar ideas [@andrews2006; @HagenEtAl2008; @nock2008; @Stengel1956] suggest that depression symptoms that put one's value to others at risk, such as suicidality and loss of interest, might be credible signals of need.

@@

## Depression and suicide attempts as both signals and bargaining tools

@@

The current study is a refinement of a large MTurk experimental vignette pilot study (N=`r nrow(signalingdata2018b)`) that used a different vignette but very similar signals and outcomes, and a much smaller pilot study posted to [reddit.com/r/SampleSize/](https://www.reddit.com/r/SampleSize/) (N=28) that used draft versions of three vignettes used in the current study, along with the same signals and outcomes. One major goal of the MTurk pilot study was to determine if believability and willingness to help were simply artifacts of the fictional victim's psychiatric distress. We therefore included a "signaling" condition in which the victim exhibited schizophrenic symptoms. As predicted, believability and willingness to help in this condition were dramatically lower than in any other condition (see Figure \@ref(fig:piloteffects)), ruling out this alternative explanation. We therefore did not include the schizophrenic condition in the current study. See the SI for more details on the pilot studies.

@@

We used the pilot data to estimate the sample sizes needed to detect an effect of the *Mild depression* signal vs. *Verbal request* control on *Belief* in the victim's need. Power was about 80% for a sample size of about `r pwr80_90[[1, 1]]`, and was about 90% for a sample size of about `r pwr80_90[[1, 2]]`. See Figure \@ref(fig:power). Given our \$1500 USD budget, we aimed for a sample size of 120-130 for treatment plus control conditions, and 1200-1300 for all conditions in the study. For more details, see the SI.

@@

Participants for this study were recruited from Amazon Mechanical Turk, a crowdsourcing platform that allows for the creation of Human Intelligence Tasks (HITs) that workers can complete for pay. As Amazon provides the infrastructure, it allows for a relatively low-cost way of collecting data for academic research at the cost of the data not being representative of any real population [@thomas2017]. Despite this limitation, it is more diverse than most convenience samples [@kennedy2020; @DworkinEtAl2016; @thomas2017], with Turkers in the US mainly differing in being younger, more educated, and lower income than the greater U.S. population [@RossEtAl2010; @boas2020].

@@

We therefore fit models of our two outcome variables (which we rescaled to the continuous interval [0, 1]) in five ways. First, we fit the preregistered OLS models (reported in the SI). Second, we fit fractional regression models, introduced in a seminal paper by @papke1996 for continuous outcomes on the interval [0, 1], which, in brief, involve fitting a quasi-likelihood version of logistic regression with robust standard errors. This method is implemented in the `frm` package [@ramalho2019]. Third, we fit GLM logistic regression models, computing 95% CI's via bootstrapping, using the `glmmboot` package created to model continuous outcomes on [0, 1] [@humphrey2020]. Fourth, we fit GLM models with the quasibinomial family and logit link using the `glm` function in R stats library, which can fit continuous outcomes on [0, 1]. Finally, we fit the quasibinomial models with robust standard errors. The estimated coefficients, standard errors, and confidence intervals produced by the latter four methods were *very* similar, and often identical (see Figure \@ref(fig:modelcompareplots). We therefore proceeded to fit and report models using `glm` with the quasibinomial family and logit link because model objects produced by this function are compatible with numerous other R packages for summarizing and plotting effects.

@@

This study was funded through a Washington State University, Vancouver research mini grant. We thank Anne Pisor, Darcy Bird, and members of the Hagen lab group for numerous helpful comments. We also thank XX and YY [pending permission to acknowledge] for feedback on the Thwarted marriage vignette.

# To do:

-   victim or signaler?
-   add anger
-   Intro "State the objectives of the work and provide an adequate background, avoiding a detailed literature survey or a summary of the results."
-   Discussion " This should explore the significance of the results of the work, not repeat them. A combined Results and Discussion section is often appropriate. Avoid extensive citations and discussion of published literature."
-   mention asking for prefered sex

# Notes from my meeting with Anne

-   Table detailing the cooperative endeavor (maybe word as boundedness of fate), conflict of interest, and what the private information is about would be useful.
-   Calling verbal request control when it's a treatment?
-   Can shift much of the pilot studies info to SI with some detail post line 279 in draft she saw.
-   Cumulative distribution plot: needs more info in captions. could benefit from fig6a and fig6b (confusion over x and y axes)

@@

## Pilot study descriptions

Briefly, participants read a vignette that put them in situations in which an imagined sister was requesting a loan of \$50,000 from the participant that was originally saved for the participant's imagined daughter's college education. The sister claims the money is needed for medical treatment for her own child, but she has incentives to deceive. The manipulations were signals very similar to those used in this study, and the outcomes were belief that the sister really needed the money and intention to loan her the money. See XX for the full write-up. See <https://osf.io/3rg8b> for preregistration of the pilot study. See <http://doi.org/10.5281/zenodo.4637883> for pilot data.

@@

## Lessons learned from the pilot study

@@

## Power from pilot study

We estimated power using our pilot study data to compare the effect of the *Mild depression* signal vs. the *Verbal request* control on *Belief* (that the sister needed the money) at T2 controlling for *Belief* at T1 in a linear regression model. Specifically, we resampled the pilot data with replacement, fit the linear regression model, and extracted the p-value for the signal coefficient, repeating this 2000 times for each sample size ranging from N=20 to N=200. Power was computed as the proportion of p-values \< 0.05. Power was about 80% for a sample size of about `r pwr80_90[[1,1]]`, and was about 90% for a sample size of about `r pwr80_90[[1,2]]`. See Figure \@ref(fig:power).

@@

## Vignettes (main bodies and signals)

>>>>

---
title: "Depression and suicidality as credible signals of need"
author: "Michael R. Gaffney, Kai Adams, Kristen L. Syme, and Edward H. Hagen"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: false
output: 
  bookdown::html_document2:
    number_sections: false
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

Major depression and suicide are generally viewed as pathological responses to stress that elicit aversive responses from others. Alternatively, depression and suicide attempts might involve honest signals of need in response to stress and help to bargain for increased support from reluctant others. To address this hypothesis, we examined responses to emotional signals in an experimental vignette study involving claims of substantial need in the presence of conflicts of interest and private information about whether the signaler was actually in need. In our sample of 1,240 participants recruited from Amazon Mechanical Turk, costlier signals like depression and suicide attempts resulted in greater perceptions of need, reduced perceptions of manipulativeness, and increased likelihood of support compared to simple verbal requests and crying without further symptoms. Additionally, as predicted, the effect of signal type on likelihood of support was largely mediated by the effect of signaling on belief in need. Our results suggest that depression and suicide attempts, apparent human universals, are relatively successful ways of signaling honest need and eliciting support, consistent with costly signaling explanations.

@@

With a cooperative species like our own, the ubiquity of conflicts of interests means that there will always be disagreement over the levels of investment in a cooperative endeavor and the division of the resulting benefits. In many such cases, individuals who were dissatisfied with the terms of cooperation could seek other partners [@hammerstein2016]. In many other cases, however, it would be difficult or impossible to switch partners. Spouses who were dissatisfied with their partner's investment in their new infant, for example, would not easily find a different partner to invest in that infant [@Hagen:1999]. Similarly, an adolescent who was dissatisfied with her parent's investment in her would not easily find other parents who were willing to invest more, nor could parents easily produce another adolescent. In these examples, and many cooperative endeavors central to human biological fitness, each party has monopoly power over the benefits they bring to the endeavor -- no one is easily replaced [@Hagen2002; @Hagen2003]. Such interdependence is increasingly recognized to be important to the evolution of cooperation in humans and other animals [@roberts2005; @tomasello2012; @aktipis2018].

@@

Formal models of bargaining highlight the role of private information regarding the value each party places the cooperative endeavor. Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay* -- the more one values the endeavor, the greater the cost of delay. In a classic non-cooperative game theory model of bargaining over the division of benefits, @rubinstein1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay. If valuations are private information, however, then immediate agreement might not be possible because each party has an incentive to request an amount greater than their actual valuation, which would harm the other party. A willingness to delay is a credible signal that one has a low valuation of the endeavor, and is therefore genuinely in need, because for those with a low valuation the benefit of waiting for a better offer outweighs the low cost of delay whereas for those for those with a high valuation the benefit of waiting for a better offer does not outweigh the higher cost of delay. Once valuations are known, the game reduces to the one in @rubinstein1982, and the parties can reach an agreement on the division of benefits [@KennanWilson1993; @feinberg2005]. @Hagen2003 proposed that the substantial reduction in productivity in depression corresponds to a willingness to delay, and is therefore a credible signal of low valuation and need. In other words, reduced productivity is relatively less costly for those whose efforts are currently not yielding many fitness-relevant benefits than for those whose efforts are yielding substantial benefits [@Hagen2003].

@@

Suicidality is closely linked to depression. Suicidal ideation is one of nine diagnostic criteria for a major depressive episode (MDE) [@dsm5], it is associated with depression across cultures [@haroz2017], and depression is a major risk factor for suicidal behavior [@hawton2013].

Anthropology has long viewed suicidality as largely the result of social problems. Early in the field's history, anthropologists tackled the problem of suicide as manifested in the small, traditional societies in which they conducted their fieldwork. Suicide, they found, was commonly a form of protest, revenge, and/or appeal [@firth1936; @Firth:1961; @Malinowski:1932; @niehaus2012]. Some ethnographers conceptualized suicide as a form of anger or social pressure [@Hezel:1987; @Giddens:1964], whereas others emphasized the powerlessness of suicide victims [@Counts:1980].

Common to almost all theoretical and empirical investigations of suicide is a focus on *completed* suicides, i.e., suicide deaths. The vast majority of suicidal behavior, however, does not result in death. In young adult women in the US, for example, there are hundreds of attempts for every death (see Figure \@ref(fig:cdc)). @SymeEtAl2016 therefore argued that the theoretical focus should be on suicide ideation and suicide attempts.

Raymond Firth was one of the first to view suicidality as a gamble to improve one's circumstances in the here and now. Based on observations that suicide attempts often followed loss or conflict and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, or immediate reintegration into the community following negative events [@firth1936; @Firth:1961].

<!-- @Rosenthal1993 proposed a credible signaling model of what he, like Firth, viewed as 'gambling-type' suicide attempts. In the Rosenthal model, as long as that risk is greater than or equal to the net benefit a non-needy person would realize by sending the signal (discounted by the risk of death), social partners can trust that signal senders are genuinely in need despite conflicts of interest [@Rosenthal1993].  -->

In the bargaining framework, suicidality is conceptualized as putting all future fitness benefits at risk with some low but non-zero probability, credibly signaling low valuation of current circumstances [see @Rosenthal1993 for a similar game theory model of suicide attempts as costly signals]. On this view, most suicides deaths, especially in young, physically healthy individuals, would therefore be the inevitable consequence that some individuals lose this gamble. These and similar ideas [@andrews2006; @HagenEtAl2008; @nock2008; @Stengel1956] suggest that depression symptoms that put one's value to others at risk, such as loss of interest and suicidality, are credible signals of need.

@@

@@

The current study is a refinement of a large MTurk experimental vignette pilot study (N=`r nrow(signalingdata2018b)`) that used a different vignette but very similar signals and outcomes (see below for more details on MTurk samples), and a much smaller pilot study posted to [reddit.com/r/SampleSize/](https://www.reddit.com/r/SampleSize/) (N=28) that used draft versions of three vignettes used in the current study, along with the same signals and outcomes. One major goal of the MTurk pilot study was to determine if believability and willingness to help were simply artifacts of the fictional victim's psychiatric distress. We therefore included a "signaling" condition in which the victim exhibited schizophrenic symptoms. As predicted, believability and willingness to help in this condition were dramatically lower than in any other condition (see Figure \@ref(fig:piloteffects)), ruling out this alternative explanation. We therefore did not include the schizophrenic condition in the current study. See the SI for more details on the pilot studies.

@@

We used the MTurk pilot data to estimate the sample sizes needed to detect an effect of the *Mild depression* signal vs. *Verbal request* control on *Belief* in the victim's need. Power was about 80% for a sample size of about `r pwr80_90[[1, 1]]`, and was about 90% for a sample size of about `r pwr80_90[[1, 2]]`. See Figure \@ref(fig:power). Given our \$1500 USD budget, we aimed for a sample size of 120-130 for treatment plus control conditions, and 1200-1300 for all conditions in the study. For more details, see the SI.

@@

Participants for this study were recruited from Amazon Mechanical Turk, a crowdsourcing platform that allows for the creation of Human Intelligence Tasks (HITs) that workers can complete for pay. As Amazon provides the infrastructure, it allows for a relatively low-cost way of collecting data for academic research at the cost of the data not being representative of any real population [@thomas2017]. Despite this limitation, MTurk samples have a wider range of ages and incomes than most university samples [@kennedy2020; @DworkinEtAl2016; @thomas2017]. US MTurk samples also differ from the general US population, mainly in being younger, more educated, and lower income [@RossEtAl2010; @boas2020].

@@

We therefore fit models of our two outcome variables (which we rescaled to the continuous interval [0, 1]) in five ways. First, we fit the preregistered OLS models (reported in the SI). Second, we fit fractional regression models, introduced in a seminal paper by @papke1996 for continuous outcomes on the interval [0, 1], which, in brief, involve fitting a quasi-likelihood version of logistic regression with robust standard errors. This method is implemented in the `frm` package [@ramalho2019]. Third, we fit GLM logistic regression models, computing 95% CI's via bootstrapping, using the `glmmboot` package created to model continuous outcomes on [0, 1] [@humphrey2020]. Fourth, we fit GLM models with the quasibinomial family and logit link using the `glm` function in R stats library, which can fit continuous outcomes on [0, 1]. Finally, we computed robust standard errors and 95% confidence intervals from the quasibinomial models using the `sandwich` package [@zeileis2006]. The estimated coefficients, standard errors, and confidence intervals produced by the latter four methods were *very* similar, and often identical (see Figure \@ref(fig:modelcompareplots). We therefore proceeded to fit and report models using `glm` with the quasibinomial family and logit link because model objects produced by this function are compatible with numerous other R packages for summarizing and plotting effects.

@@

This study was funded through a Washington State University, Vancouver Research Mini Grant. We thank Anne Pisor, Darcy Bird, and members of the Hagen lab group for numerous helpful comments. We also thank XX and YY [pending permission to acknowledge] for feedback on the Thwarted marriage vignette.

<!-- # To do: -->

<!-- -   victim or signaler? -->
<!-- -   add anger -->
<!-- -   Intro "State the objectives of the work and provide an adequate background, avoiding a detailed literature survey or a summary of the results." -->
<!-- -   Discussion " This should explore the significance of the results of the work, not repeat them. A combined Results and Discussion section is often appropriate. Avoid extensive citations and discussion of published literature." -->
<!-- -   mention asking for prefered sex -->

<!-- # Notes from my meeting with Anne -->

<!-- -   Table detailing the cooperative endeavor (maybe word as boundedness of fate), conflict of interest, and what the private information is about would be useful. -->
<!-- -   Calling verbal request control when it's a treatment? -->
<!-- -   Can shift much of the pilot studies info to SI with some detail post line 279 in draft she saw. -->
<!-- -   Cumulative distribution plot: needs more info in captions. could benefit from fig6a and fig6b (confusion over x and y axes) -->

@@

## Pilot studies

We conducted two pilot studies, one on MTurk with a large sample size (N=`r `nrow(signalingdata2018b)`), and one on [reddit.com/r/SampleSize](https://reddit.com/r/SampleSize) with a small sample size (N=28). Both studies used the experimental pre-test, post-test design used in the current study.

In the MTurk study, participants read a vignette that put them in situations in which an imagined sister was requesting a loan of \$50,000 from the participant that was originally saved for the participant's imagined daughter's college education. The sister claims the money is needed for medical treatment for her own child, but she has incentives to deceive. The manipulations were signals very similar to those used in this study, and the outcomes were belief that the sister really needed the money and intention to loan her the money. See XX for the full write-up. See <https://osf.io/3rg8b> for preregistration of the pilot study. See <http://doi.org/10.5281/zenodo.4637883> for pilot data.

@@

@@

## Power from MTurk pilot study

We estimated power using our MTurk pilot study data to compare the effect of the *Mild depression* signal vs. the *Verbal request* control on *Belief* (that the sister needed the money) at T2 controlling for *Belief* at T1 in a linear regression model. Specifically, we resampled the pilot data with replacement, fit the linear regression model, and extracted the p-value for the signal coefficient, repeating this 2000 times for each sample size ranging from N=20 to N=200. Power was computed as the proportion of p-values \< 0.05. Power was about 80% for a sample size of about `r pwr80_90[[1,1]]`, and was about 90% for a sample size of about `r pwr80_90[[1,2]]`. See Figure \@ref(fig:power).

@@

## Vignettes of the current study (main bodies and signals)
